Improving the portuguese hyphenation rules

### 2024-04-18 article/article.tex
updates

\usepackage[textsize=tiny,textwidth=3.5cm,color=blue!10]{todonotes}

@@

@@

Each language has its hyphenation rules, which can be categorized into two
groups: those driven by morphology (etymology) and those driven by
pronunciation.  An algorithmic approach employs a logical system to analyze
words and apply the hyphenation rules of a specific language. Since hyphenation
rules vary significantly between languages, a unique algorithm must be
developed for each one. While a logic-based system can be efficient and
compact, it still needs to address exceptions through hard-coded rules.

@@

Unfortunately, certain hyphenation rules cannot be implemented using the \TeX{}
hyphenation algorithm. In languages like German, hyphenation can lead to letter
change or insertion. Additionally, compound words lack hyphens, resulting in
extended letter sequences without visible separation and even repetitions of
the same letter, as seen in examples like \emph{Wasserrinne} and
\emph{Schifffahrt}.  Furthermore, the German spelling reform made some changes,
making it necessary to create a different set of rules for German hyphenation.
For example, the word \emph{Schiffahrt} should be hyphenated as
\emph{Schiff-fahrt}, preserving the \emph{f}s from each word that makes this
compound. The hyphenation should insert an \emph{f} that is not part of the
written form. That was not a problem for the old written form of the word:
\emph{Schifffahrt}. Also, the old hyphenating rules of German grammar stated
the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker}, \emph{Zuk-ker} for
the word \emph{Zucker} and \emph{pak-ken} for the word \emph{packen}. Now those
words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker} and \emph{pa-cken},
respectively. Some words have also changed their hyphenation point after the
spelling reform. For example, \emph{Fen-ster} became \emph{Fens-ter} and
\emph{mei-stens} became \emph{meis-tens}. Some problems in compound word
hyphenation in \TeX{} are discussed in \textcite{sojka1995a}.

@@

To simplify, if we consider only the Latin alphabet, with no diacritcs, the patterns used 
in \TeX{} hyphenation are of the form: \texttt{\^{}\textbackslash{}.?[0-9]?([a-z]+[0-9]?)+\textbackslash{}.?\$},
where we have described it using a regular expression\footnote{
    Regular expressions (regex) are powerful search patterns used in text processing to find, match, and manipulate strings of text. They are a fundamental tool in programming and are supported in many programming languages.
    The regex given here uses the Perl syntax and might be broken in the following
    parts: \texttt{\^{}} and \texttt{\$} mark the start and end of the string;
    \texttt{\textbackslash{}.?} specifies an optional period; \texttt{[0-9]?} allows for an optional single digit;
    \texttt{([a-z]+[0-9]?)+} matches sequences of lowercase letters interleaved with optional digits.
}.
One example of such pattern is \texttt{4z1z2}, which is composed of a sequence
of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. Patterns may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.sh2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

Hyphenation rules are organized in levels, from 1 to 9, where odd numbers
represent hyphenating levels and even numbers represent inhibiting levels. Each
level works as an exception level of it predecessor. For example, the rule
\texttt{sh1er} indicate a good hyphenation point between the \emph{h} and the
\emph{e} in the sequence \emph{sher}. A rule at a higher level, as
\texttt{.sh2}, implies an exception to the lower level rule. When we see
\emph{sher}, in the beginning of a word, the rule \texttt{.sh2} applies and
hyphenation proposed by the lower level rule \texttt{sh1er} should be hindered.
That is the case in the hyphenation of the word \emph{Sher-lock}. The full
example is provided in \Cref{sherlockhyphenation}, where we might see all pertinent
English rules taking place in the hyphenation of \emph{Sher-lock} and \emph{sher-iff}.

@@

A pattern will consist of a string made of characters (from the language
alphabet) possibly with a number in between, expressing the
hyphenation/inhibition level and possibly word boundaries marker (the period)
at the pattern edges. When there is no number between characters in a pattern,
a zero is assumed, which means \emph{undefined} and no hyphenation point will
be suggest at that location.

@@

PATGEN utilizes a list of hyphenated words to define rules at various levels
and lengths, based on the patterns found in the data. It starts with short
patterns and incrementally increases their length until reaching the maximum
pattern length allowed by the user.  The objective is to keep the patterns as
concise as possible, as this enhances their generalizability.  As it advances
and incorporates longer patterns, PATGEN establishes exceptions.  In certain
cases, analyzing long patterns may be necessary, as some hyphenation points
could depend on characters far away from the breaking point\footnote{Some
    examples of hyphenation dependency on characters far from the break point:
    \emph{dem-o-crat} and \emph{de-moc-ra-cy}; \emph{as-pi-rin} and
    \emph{aspir-ing}; \emph{de-mon-stra-tive} and \emph{dem-on-stra-tion}.
}.

PATEGEN works on glyph\footnote{\emph{Glyph} is commonly used in linguistics,
typography, and computer graphics to refer to a specific graphical
representation of a character or symbol, which can be the entire symbol or a
distinct visual element within it.} indices rather than character codes. Each
glyph is represented by a single byte. That amount to 256 indices, where 13 of
them are reserved for the digits 0-9 and the characters `.', `-', and `*'. The
remaining 243 are used to represent symbols of the given language.  To run
PATGEN, a translation file is necessary. This file defines the values of
certain language-specific parameters (in the first line) and enumerates the
various forms in which language symbols may appear (all subsequent lines).  In
the first line, positions 1 and 2 are used to set the value of
\verb|lefthyphenmin|, and positions 3 and 4 are used to set the value of
\verb|righthyphenmin|.  These values determine the minimum length of a string
that may be generated by a hyphenation procedure. To set a single-digit value,
leave the first position blank, i.e., place a space in position 1 and 3 for
\verb|lefthyphenmin| and \verb|righthyphenmin|, respectively.  Positions 5, 6
and 7 are used to define alternative values for the special characters `.', `-'
and `*'.  The following lines use a delimiter to enclose each `letter' of the
desired language alphabet, including its alternative representations.  The
first position of the line defines the delimiter, and each symbol of the
language can occupy as many positions as necessary, as long as the reserved
value for the delimiter is not used in the symbol's definition.  Consider the
following example for defining the letter `e' in Portuguese: 
\begin{verbatim}
XeXEXéXêXÉXÊX\'{e}X\^{e}X\'{E}X\^{E}XX 
\end{verbatim}
It represents the many ways in which the letter `e' might be found: lowercase,
uppercase, with or without acute or circumflex accents. Note that we have used
the direct input (using UTF-8 or other encoding that support the accented `e'
character) and also the compositional counterpart using the escape codes that
define each accent.  In this definition, we have assumed that the many forms in
which we may find the character \emph{e} will be equivalent for
hyphenation (pattern matching) purpose. As another example, see the next line
which defines the character $\pi$ (taken from \textcite{haralambous2021}):
\begin{verbatim} 
#p#P#\varpi ## 
\end{verbatim}

@@

%
% FILE="../data/default.TeX.pt-br.patterns"; grep "^1..$" "$FILE" | awk '{print substr($0, 0, 2)}' | sort | uniq -c | while read line; do p=$(echo $line | cut -d' ' -f2); c=$(grep "^$p[^0-9]$" "$FILE" | awk '{print substr($0, 3)}' | tr '\n' ' '); printf "$line $c \n"; done | cut -d' ' -f1 | paste -sd+ | bc
%
Out of the default \TeX{} rules, 252 (82\%) follow the pattern \verb|1CV|, which
represents the recurring CV syllables in Portuguese.  As for the consonants,
there are typically 18 considered (\verb|b|, \verb|c|, \verb|ç|, \verb|d|,
\verb|f|, \verb|g|, \verb|j|, \verb|k|, \verb|l|, \verb|m|, \verb|n|, \verb|p|,
\verb|r|, \verb|s|, \verb|t|, \verb|v|, \verb|x|, and \verb|z|) that can
combine with 14 vowels (\verb|a|, \verb|e|, \verb|i|, \verb|o|, \verb|u|,
\verb|á|, \verb|â|, \verb|ã|, \verb|é|, \verb|í|, \verb|ó|, \verb|ú|, \verb|ê|,
and \verb|õ|), resulting in these CV patterns that indicate favorable
hyphenation points before the consonants\footnote{The complete list of these
252 patterns is given here: \ExceptionRulesCV{}.}. 
It is worth noting that there might
be other rules or exceptions in the hyphenation patterns not covered by these
default rules.  To accommodate exceptions, \textcite{rezende1987} proposes the 
following rules: 
\begin{itemize}
    \item 20 rules that created for cases involving consonants \verb|b|, \verb|c|, \verb|d|, \verb|f|, \verb|g|, \verb|k|, \verb|p|, \verb|t|, \verb|v|, or \verb|w| followed by \verb|l| or \verb|r|\footnote{The 20 additional rules are: \ExceptionRulesLR{}.};
    \item 3 rules for \verb|c|, \verb|l| or \verb|n| followed by \verb|h|\footnote{Those 3 rules are: \ExceptionRulesCLNH{}.}.
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep "2" | awk '{print substr($0, 2, 1)}' | sort | uniq | while read line; doc=$(grep "^1${line}2.$" "$FILE" | tr -d '12' | awk '{print substr($0, 2, 1)}' | tr '\n' ' '); printf "$line $c\n"; done
    \item 23 distinct patterns were introduced to indicate hyphenation points between vowels or between \verb|c|'s, \verb|r|'s, and \verb|s|'s\footnote{Those patterns are: \ExceptionRulesThree{}.}. 
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep -v "2" | grep "3" | wc -l
    \item 8 patterns adhere to the \verb|1[gq]u4V|, signaling beneficial hyphenation points before \verb|g| or \verb|q|, followed by a sequence of \verb|u| and a vowel, with an inhibiting point between the \verb|u| and the subsequent vowel\footnote{They are the following: \ExceptionRulesFour{}.}.
    \item 1 pattern represented as \verb|1-|, denoting that a hyphen indeed serves as a beneficial hyphenation point.
\end{itemize}

@@

\begin{table}
\centering
\caption{Considering the six dictionaries used, the first line of this table
presents the number words that a given number of hyphenations were found in the
dictionaries. The second line presents the number of words that have a given
number of hyphenation as its most frequent form found in the
dictionaries.}\label{tblstats}
\small
\begin{tabular}{@{ }l@{ } *{7}{@{ }S[table-format=6]@{ }}}
    hyphenations & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
    \hline
    hyphenations found & \NumberOfSixHyphens{} & \NumberOfFiveHyphens{} &
    \NumberOfFourHyphen{} & \NumberOfThreeHyphens{} & \NumberOfTwoHyphens{} &
    \NumberOfOneHyphens{} & \NumberOfNoHyphens{} \\

@@

There is a diverse array of writing systems, categorized into logographic,
syllabic, and alphabetic systems. Although distinct, these systems can be built
on an interplay of these categories \parencite{coulmas2003,palmer2010}. 
The principles guiding an alphabetic or phonemic writing system vary significantly 
based on the language and its history. Here are some key points worth highlighting:
\begin{enumerate}[label=\arabic*)]
    \item The most common principle involves representing the sounds of a
	language with written symbols, using one or a combination of symbols to
	represent each sound (e.g., the orthographic system of Portuguese uses
	an alphabet of 27 letters to represent different sounds of the
	language).
    \item Etymology also plays a crucial role, where the spelling of a word
	reflects its origins and historical development (e.g., the French
	language often reflects the Latin roots of words in its spelling).
    \item Morphology serves as a guide in structuring many languages, using
	specific letters or symbols to indicate word endings, prefixes, or
	suffixes (e.g., Russian uses different forms of the Cyrillic alphabet
	to indicate gender and case in its nouns).
    \item The evolution of a language over time also contributes to its written
	form (e.g., the spelling of many English words has changed over time to
	reflect changes in pronunciation).
\end{enumerate}
Orthographic systems for languages with logographic writing systems, such as
Chinese ideograms, cuneiform writing, and Egyptian hieroglyphs, differ
significantly from alphabetic writing systems. In logographic systems,
individual symbols or characters represent entire words or ideas, rather than
phonemes or sounds. As a result, the principles guiding their orthographic
systems are based more on semantic and visual principles than on phonemic
principles.
Hyphenation is a feature of alphabetic scripts where words are composed of letters, 
and spaces or hyphens are used to separate words, syllables, or parts of words. 
Logographic writing systems, such as the Chinese, 
function differently and do not employ hyphenation \parencite{honorof2006}.

@@

Portuguese employs an alphabetical writing system, which means its spelling is
guided by phonological principles \parencite{cagliari2015}. The correlation between
spelling and pronunciation influences word hyphenation, as words are divided
into syllables based on the phonemic system. It is important to note that
different languages follow diverse principles for word division. For example,
English is primarily guided by morphological principles, evident in words like
\emph{walk-ing}, \emph{un-happy}, \emph{work-s}, and \emph{ear-ly}.
Additionally, other factors also impact word hyphenation in English, such as 
the distinction between long and short vowels, which function within the context
of open or closed syllable, respectively; and the presence of doubled consonants and
digraphs \parencite{lin2011,yavas2020}. 
Furthermore, there exist numerous exceptions that may be categorized into rules.
% https://english.stackexchange.com/questions/385/what-are-the-rules-for-splitting-words-at-the-end-of-a-line

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly mirror the
phonetic counterpart. This is notably apparent in Portuguese, where a strict
one-to-one correspondence between letters and sounds is not always observed.
The orthographic system operates according to its specific rules. For example,
consider consonant clusters that create a single sound (digraphs) in words like
\emph{\underline{ch}ave}, \emph{i\underline{lh}a}, \emph{\underline{sh}ampoo}, \emph{ca\underline{rr}o}, and \emph{ma\underline{ss}a}.
While these digraphs are pronounced as one sound within a single syllable,
their representation in writing determines how they are divided. Specifically,
different consonants within a digraph must remain together, whereas identical
consonants are separated. As a result, we observe hyphenations like
\emph{cha-ve}, \emph{i-lha}, and \emph{sham-poo}, but \emph{car-ro} and \emph{mas-sa}.

@@

Systematizing the rules that guide syllable boundaries and, consequently,
hyphenation in Portuguese proved to be fundamental for the improvement of the
\TeX{} hyphenator rules, since they indicate the possible or prohibited contexts
of hyphenation. Expomos, a seguir, uma análise comparativa...

... to continue
\vspace{10ex}

\section{Creating the dictionary}

@@

After performing all procedures above described to create our golden hyphenation
dictionary, we might still need manual corrections.
All manual corrections are written in the file \texttt{replacements.txt} and
the script \texttt{manualcorrections.sh} is used to perform these corrections.

>>>>

%\usepackage[textsize=tiny,textwidth=3.5cm,color=blue!10]{todonotes}

@@

\usepackage{rotating}
\usepackage{bytefield}

@@

Each language has its hyphenation rules, which can be categorized into two
groups: those driven by morphology (etymology) and those driven by
pronunciation.  An algorithmic approach employs a logical system to analyze
words and apply the hyphenation rules of a specific language. Since hyphenation
rules vary significantly between languages, an algorithm must be
developed for each one. While a logic-based system can be efficient and
compact, it still needs to address exceptions through hard-coded rules.

@@

Unfortunately, certain hyphenation rules cannot be implemented using the \TeX{}
hyphenation algorithm. For example, in German, hyphenation can lead to letter
change or insertion.  Additionally, compound words lack hyphens, resulting in
extended letter sequences without visible separation and even repetitions of
the same letter, as seen in examples like \emph{Wasserrinne} and
\emph{Schifffahrt}.  Furthermore, the German spelling reform made some changes,
making it necessary to create a different set of rules for German hyphenation.
For example, the word \emph{Schiffahrt} should be hyphenated as
\emph{Schiff-fahrt}, preserving the \emph{f}s from each word that makes this
compound. The hyphenation should insert an \emph{f} that is not part of the
written form. That was not a problem for the old written form of the word:
\emph{Schifffahrt}. Also, the old hyphenating rules of German grammar stated
the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker}, \emph{Zuk-ker} for
the word \emph{Zucker} and \emph{pak-ken} for the word \emph{packen}. Nowadays, those
words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker} and \emph{pa-cken},
respectively. Some words have also changed their hyphenation point after the
spelling reform. For example, \emph{Fen-ster} became \emph{Fens-ter} and
\emph{mei-stens} became \emph{meis-tens}. Some problems in compound word
hyphenation in \TeX{} are discussed in \textcite{sojka1995a}.

@@

To simplify, if we consider only the Latin alphabet, with no diacritcs, the patterns used 
in \TeX{} hyphenation are of the form: \texttt{\^{}\textbackslash{}.?[0-9]?([a-z]+[0-9]?)+\textbackslash{}.?\$},
where we have described it using a regular expression\footnote{
    Regular expressions (regex) are powerful search patterns used in text processing to find, match, and manipulate strings of text. They are a fundamental tool in programming and are supported in many programming languages.
    The regex given here uses the Perl syntax and might be broken in the following
    parts: \texttt{\^{}} and \texttt{\$} mark the start and end of the string;
    \texttt{\textbackslash{}.?} specifies an optional period; \texttt{[0-9]?} allows for an optional single digit;
    \texttt{([a-z]+[0-9]?)+} matches sequences of lowercase letters interleaved with optional digits.
}.
One example of such pattern is \texttt{4z1z2}, which is composed of a sequence
of letters and numbers. In general, we use characters/symbols from the language
alphabet, along with Hindu-Arabic numbers, to express either hyphenation
facilitation or inhibition.  Odd numbers indicate a good hyphenation point,
whereas even numbers indicate a bad place to break.  The given example states
that the sequence has a good breaking point between the first and the second
\emph{z} and an hyphenation should be inhibited before the first \emph{z} and
after the second \emph{z}. For example, the hyphenation of the word
\emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule, where we see
the hyphen placed between the two \emph{z}'s and no hyphen before, nor after
the \emph{z}'s. Patterns may also use period symbol (\emph{.}) to indicate word
boundaries. The pattern \texttt{.sh2} applies to beginning of words, implying
that the \emph{s} and the \emph{h} should stick together in beginning of a word
and an hyphenation should also be inhibited after the \emph{h}. For example,
this pattern is used in \emph{Sher-lock}.

Hyphenation rules are organized in levels, from 1 to 9, where odd numbers
represent hyphenating levels and even numbers represent inhibiting levels. Each
level works as an exception level of it predecessor. For example, the rule
\texttt{sh1er} indicate a good hyphenation point between the \emph{h} and the
\emph{e} in the sequence \emph{sher}. A rule at a higher level, as
\texttt{.sh2}, implies an exception to the lower level rule. When we see
\emph{sher}, in the beginning of a word, the rule \texttt{.sh2} applies and
hyphenation proposed by the lower level rule \texttt{sh1er} should be hindered.
That is the case in the hyphenation of the word \emph{Sher-lock}. The full
example is provided in \Cref{sherlockhyphenation}, where we might see all pertinent
English rules taking place in the hyphenation of \emph{Sher-lock}.

@@

In summary, a pattern will consist of a string made of characters (from the language
alphabet) possibly with a number in between, expressing the
hyphenation/inhibition level. Occasionally word boundaries marker (the period)
is used at the pattern edges. When there is no number between characters in a pattern,
a zero is assumed, which means \emph{undefined} and no hyphenation point will
be suggest at that location.

@@

PATGEN utilizes a list of hyphenated words to extract patterns and use them to define rules at various levels and lengths.
It starts with short
patterns and incrementally increases their length until reaching the maximum
pattern length allowed by the user.  The objective is to keep the patterns as
concise as possible, as this enhances their generalizability.  As it advances
and incorporates longer patterns, PATGEN establishes exceptions.  In certain
cases, analyzing long patterns may be necessary, as some hyphenation points
could depend on characters far away from the breaking point\footnote{Some
    examples of hyphenation dependency on characters far from the break point:
    \emph{dem-o-crat} and \emph{de-moc-ra-cy}; \emph{as-pi-rin} and
    \emph{aspir-ing}; \emph{de-mon-stra-tive} and \emph{dem-on-stra-tion}.
}.

PATEGEN works on glyph\footnote{\emph{Glyph} is commonly used in linguistics,
typography, and computer graphics to refer to a specific graphical
representation of a character or symbol, which can be the entire symbol or a
distinct visual element within it.} indices rather than character codes. Each
glyph is represented by a single byte. That amount to 256 indices, where 13 of
them are reserved for the digits 0-9 and the characters `.', `-', and `*'. The
remaining 243 are used to represent symbols of a given language.  To run
PATGEN, a translation file is necessary. This file defines the values of
certain language-specific parameters (in the first line) and enumerates the
various forms in which language symbols may appear (all subsequent lines).  In
the first line, positions 1 and 2 are used to set the value of
\verb|lefthyphenmin|, and positions 3 and 4 are used to set the value of
\verb|righthyphenmin|.  These values determine the minimum length of a string
that may be generated by a hyphenation procedure. To set a single-digit value,
leave the first position blank, i.e., place a space in position 1 and 3 for
\verb|lefthyphenmin| and \verb|righthyphenmin|, respectively.  Positions 5, 6
and 7 are used to define alternative values for the special characters `.', `-'
and `*'.  
\begin{center}
\begin{bytefield}[endianness=little,bitwidth=0.11111\linewidth,boxformatting={\centering}]{7}
\bitheader[lsb=1]{1-7} \\
\bitbox{2}{lefthyphenmin} & \bitbox{2}{righthyphenmin} & \bitbox{1}{.} & \bitbox{1}{-} & \bitbox{1}{*} \\
\end{bytefield}
\end{center}
The following lines use a delimiter to enclose each `letter' of the
desired language alphabet, including its alternative representations.  The
first position of the line defines the delimiter, and each symbol of the
language can occupy as many positions as necessary, as long as the reserved
value for the delimiter is not used in the symbol's definition. The defining lines ends
when the delimiter appears twice in a roll. Consider the
following example for defining the letter `e' in Portuguese:
\verb|XeXEXéXêXÉXÊX\'{e}X\^{e}X\'{E}X\^{E}XX|.
\newcommand{\bitlabel}[2]{%
    \bitbox[]{#1}{%
	\raisebox{0pt}[4ex][0pt]{%
	    \turnbox{45}{\fontsize{7}{7}\selectfont#2}%
	}%
    }%
}
\vspace{5ex}
\begin{center}
\begin{bytefield}[bitwidth=1.9em,boxformatting={\centering\tiny}]{20}
\bitlabel{1}{delimiter} & \bitlabel{1}{letter e} &
\bitlabel{1}{delimiter} & \bitlabel{1}{letter E} &
\bitlabel{1}{delimiter} & \bitlabel{1}{letter é} &
\bitlabel{1}{delimiter} & \bitlabel{1}{letter ê} &
\bitlabel{1}{delimiter} & \bitlabel{1}{letter É} &
\bitlabel{1}{delimiter} & \bitlabel{1}{letter Ê} &
\bitlabel{1}{delimiter} & \bitlabel{1}{TeX sequence for é} &
\bitlabel{1}{delimiter} & \bitlabel{1}{TeX sequence for ê} &
\bitlabel{1}{delimiter} & \bitlabel{1}{TeX sequence for É} &
\bitlabel{1}{delimiter} & \bitlabel{1}{TeX sequence for Ê} \\
\bitheader[lsb=1]{1-20} \\
\bitbox{1}{X} & \bitbox{1}{e} & \bitbox{1}{X} & \bitbox{1}{E} &
\bitbox{1}{X} & \bitbox{1}{é} & \bitbox{1}{X} & \bitbox{1}{ê} &
\bitbox{1}{X} & \bitbox{1}{É} & \bitbox{1}{X} & \bitbox{1}{Ê} &
\bitbox{1}{X} & \bitbox{1}{\textbackslash{}'\{e\}} & \bitbox{1}{X} & \bitbox{1}{\textbackslash{}\^{}\{e\}} &
\bitbox{1}{X} & \bitbox{1}{\textbackslash{}'\{E\}} & \bitbox{1}{X} & \bitbox{1}{\textbackslash{}\^{}\{E\}}
\end{bytefield}
\end{center}
%\begin{verbatim}
%XeXEXéXêXÉXÊX\'{e}X\^{e}X\'{E}X\^{E}XX 
%\end{verbatim}
We have adopted \verb|X| as the delimiter. This line represents the many ways in which the letter `e' might be found: lowercase,
uppercase, with or without acute or circumflex accents. Note that we have used
the direct input (using UTF-8 or other encoding that support the accented `e'
character) and also the compositional counterpart using the appropriate \TeX{} control sequence that
instructs \TeX{} to place accent on the letter.  In this definition, we have assumed that the many forms in
which we may find the character \emph{e} will be equivalent for
hyphenation (pattern matching) purpose. As another example, see the next line
which defines the character $\pi$ (taken from \textcite{haralambous2021}):
\verb|#p#P#\varpi ##|.
%\begin{verbatim} 
%#p#P#\varpi ## 
%\end{verbatim}

@@

%
% FILE="../data/default.TeX.pt-br.patterns"; grep "^1..$" "$FILE" | awk '{print substr($0, 0, 2)}' | sort | uniq -c | while read line; do p=$(echo $line | cut -d' ' -f2); c=$(grep "^$p[^0-9]$" "$FILE" | awk '{print substr($0, 3)}' | tr '\n' ' '); printf "$line $c \n"; done | cut -d' ' -f1 | paste -sd+ | bc
%
Out of the default \TeX{} rules, 252 (82\%) follow the pattern \verb|1CV|, which
represents the recurring CV syllables in Portuguese.  As for the consonants,
there are typically 18 considered (\verb|b|, \verb|c|, \verb|ç|, \verb|d|,
\verb|f|, \verb|g|, \verb|j|, \verb|k|, \verb|l|, \verb|m|, \verb|n|, \verb|p|,
\verb|r|, \verb|s|, \verb|t|, \verb|v|, \verb|x|, and \verb|z|) that can
combine with 14 vowels (\verb|a|, \verb|e|, \verb|i|, \verb|o|, \verb|u|,
\verb|á|, \verb|â|, \verb|ã|, \verb|é|, \verb|í|, \verb|ó|, \verb|ú|, \verb|ê|,
and \verb|õ|), resulting in these CV patterns that indicate favorable
hyphenation points before the consonants\footnote{The complete list of these
252 patterns is given here: \ExceptionRulesCV{}.}. 
It is worth noting that there might
be other rules or exceptions in the hyphenation patterns not covered by these
default rules.  To accommodate exceptions, \textcite{rezende1987} proposes the 
following rules: 
\begin{itemize}
    \item 20 rules created for cases involving consonants \verb|b|, \verb|c|, \verb|d|, \verb|f|, \verb|g|, \verb|k|, \verb|p|, \verb|t|, \verb|v|, or \verb|w| followed by \verb|l| or \verb|r|\footnote{The 20 additional rules are: \ExceptionRulesLR{}.};
    \item 3 rules for \verb|c|, \verb|l| or \verb|n| followed by \verb|h|\footnote{Those 3 rules are: \ExceptionRulesCLNH{}.};
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep "2" | awk '{print substr($0, 2, 1)}' | sort | uniq | while read line; doc=$(grep "^1${line}2.$" "$FILE" | tr -d '12' | awk '{print substr($0, 2, 1)}' | tr '\n' ' '); printf "$line $c\n"; done
    \item 23 distinct patterns were introduced to indicate hyphenation points between vowels or between \verb|c|'s, \verb|r|'s, and \verb|s|'s\footnote{Those patterns are: \ExceptionRulesThree{}.}; 
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep -v "2" | grep "3" | wc -l
    \item 8 patterns adhere to the \verb|1[gq]u4V| pattern, signaling beneficial hyphenation points before \verb|g| or \verb|q|, followed by a sequence of \verb|u| and a vowel, with an inhibiting point between the \verb|u| and the subsequent vowel\footnote{They are the following: \ExceptionRulesFour{}.};
    \item 1 pattern represented as \verb|1-|, denoting that a hyphen indeed serves as a beneficial hyphenation point.
\end{itemize}

@@

\begin{table}
\centering
\caption{Considering the six dictionaries used, the first line of this table
presents the number of words that a given number of hyphenations were found in the
dictionaries. The second line presents the number of words that have a given
number of hyphenation as its most frequent form found in the
dictionaries.}\label{tblstats}
\small
\begin{tabular}{@{ }l@{ } *{7}{@{ }S[table-format=6]@{ }}}
    hyphenations & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
    \hline
    hyphenations found & \NumberOfSixHyphens{} & \NumberOfFiveHyphens{} &
    \NumberOfFourHyphen{} & \NumberOfThreeHyphens{} & \NumberOfTwoHyphens{} &
    \NumberOfOneHyphens{} & \NumberOfNoHyphens{} \\

@@

There is a diverse collection of writing systems, categorized into logographic,
syllabic, and alphabetic systems. Although distinct, these systems can be built
on an interplay of these categories \parencite{coulmas2003,palmer2010}. 
The principles guiding an alphabetic or phonemic writing system vary significantly 
based on the language and its history. Here are some key points worth highlighting:
\begin{enumerate}[label=\arabic*)]
    \item The most common principle involves representing the sounds of a
	language with written symbols, using one or a combination of symbols to
	represent each sound (e.g., the orthographic system of Portuguese uses
	an alphabet of 27 letters to represent different sounds of the
	language).
    \item Etymology also plays a crucial role, where the spelling of a word
	reflects its origins and historical development (e.g., the French
	language often reflects the Latin roots of words in its spelling).
    \item Morphology serves as a guide in structuring many languages, using
	specific letters or symbols to indicate word endings, prefixes, or
	suffixes (e.g., Russian uses different forms of the Cyrillic alphabet
	to indicate gender and case in its nouns).
    \item The evolution of a language over time also contributes to its written
	form (e.g., the spelling of many English words has changed over time to
	reflect changes in pronunciation).
\end{enumerate}
Orthographic systems for languages with logographic writing systems, such as
Chinese ideograms, cuneiform writing, and Egyptian hieroglyphs, differ
significantly from alphabetic writing systems. In logographic systems,
individual symbols or characters represent entire words or ideas, rather than
phonemes or sounds. As a result, the principles guiding their orthographic
systems are based more on semantic and visual principles than on phonemic
principles.
Hyphenation is a feature of alphabetic scripts where words are composed of letters, 
and spaces or hyphens are used to separate words, syllables, or parts of words. 
Logographic writing systems, such as the Chinese, 
function differently and do not employ hyphenation \parencite{honorof2006}.

@@

Portuguese employs an alphabetical writing system, which means its spelling is
guided by phonological principles \parencite{cagliari2015}. The correlation between
spelling and pronunciation influences word hyphenation, as words are divided
into syllables based on the phonemic system. It is important to note that
different languages follow diverse principles for word division. For example,
English is primarily guided by morphological principles, evident in words like
\emph{walk-ing}, \emph{un-happy}, \emph{work-s}, and \emph{ear-ly}.
Additionally, other factors also impact word hyphenation in English, such as 
the distinction between long and short vowels, which function within the context
of open or closed syllable, respectively; and the presence of doubled consonants and
digraphs \parencite{lin2011,yavas2020}.
Although each language has its driving principles for the hyphenation process, 
multiple factors come into play, leading to various solutions in certain scenarios.
For instance, in Portuguese, the phonological principle would lead to \emph{hipe-rativo}, while 
the morphological principle would lead to \emph{hiper-ativo}. Both approaches seem valid
and are indeed found in online dictionaries\footnote{Looking at words with prefixes, 
it seems that even on those words, the phonological approach was predominant on hyphenation,
but we cannot tell if it is a byproduct of automatic hyphenation that might be used
to hyphenate words on online dictionaries.}.
Some rather rare words, such as \emph{hiperalgesia}, may not be subject to morphological 
influences. Since it has a low frequency of occurrence, the individual may not be
aware of its morphological components and therefore hyphenate it as \emph{hipe-ralgesia}.
The form \emph{hiper-algesia} could also be accepted, emphasizing its morphological constituents. 
Furthermore, there exist numerous exceptions that may be categorized into rules.
% https://english.stackexchange.com/questions/385/what-are-the-rules-for-splitting-words-at-the-end-of-a-line

% $ grep "^hiper[aeiou]" ../data/hyphenations.csv  | grep "-" | grep -o "hi-pe-r[aeiou]\|hi-per-[aeiou]" | sort | uniq -c
% 16 hi-pe-ra
%  1 hi-per-a
% 29 hi-pe-re
% 20 hi-pe-ri
%  2 hi-per-i
%  2 hi-pe-ro
%  2 hi-pe-ru

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly mirror the
phonetic counterpart. This is notably apparent in Portuguese, where a strict
one-to-one correspondence between letters and sounds is not always observed.
The orthographic system operates according to its specific rules. For example,
consider consonant clusters that create a single sound (digraphs) in words like
%\emph{\underline{ch}ave}, 
\emph{a\underline{ch}ado},
\emph{i\underline{lh}a}, 
%\emph{\underline{sh}ampoo}, 
\emph{su\underline{sh}i},
\emph{ca\underline{rr}o}, and \emph{ma\underline{ss}a}.
While these digraphs are pronounced as one sound within a single syllable,
their representation in writing determines how they are divided. Specifically,
different consonants within a digraph must remain together, whereas identical
consonants are separated. As a result, we observe hyphenations like
%\emph{cha-ve}, 
\emph{a-cha-do},
\emph{i-lha}, and 
%\emph{sham-poo}, 
\emph{su-shi},
but \emph{car-ro} and \emph{mas-sa}.

@@

Systematizing the rules that guide syllable boundaries and hyphenation in
Portuguese is a fundamental step to understand and improve the \TeX{}
hyphenation rules. To ensure the accuracy of hyphenation rules and effectively
compare their results, a hyphenation spelling dictionary serves as an
indispensable reference. This dictionary provides a comprehensive listing of
correctly hyphenated words, enabling the computation of the correctness of each
set of hyphenation rules. By consulting the hyphenation spelling dictionary,
one can verify whether the hyphenation patterns generated by a set of rules
align with established standards. This process involves analyzing how well the
hyphenation rules adhere to the accepted conventions of the language, ensuring
that they effectively segment words without compromising readability or
consistency. Moreover, by comparing the results obtained from different sets of
hyphenation rules against the entries in the spelling dictionary, we can assess
the efficacy and reliability of each approach. In section \ref{sec-dictionay},
we will define the specifications of this reference dictionary, outlining its
role in subsequent rule updates and rules generation.

\section{Creating the dictionary}\label{sec-dictionay}

@@

After completing the aforementioned procedures to create our golden hyphenation
dictionary, manual corrections may still be necessary to ensure its accuracy
and reliability. These manual adjustments, which encompass the corrections
described earlier, are documented in the file \texttt{replacements.txt}.
Utilizing the script \texttt{manualcorrections.sh}, we meticulously apply these
corrections to refine our hyphenation dictionary and enhance its effectiveness.
